\toolnameend: A Tool for Visualizing Patterns of Spreadsheet Function Combinations

### 2016-05-24 doc/paper.tex
Various changes for getting the info out.

	This paper contributes such a tool, \toolname, that attempts to fulfill this
	need for better exploration by building on the aforementioned spreadsheet
	corpora. Working with spreadsheets in Excel, \toolname focuses primarily on the
	functions, like SUM and IF and hundreds more, which are the building blocks of
	formula and programs that determine values across a spreadsheet. The tool
	creates an interactive, exploratory visualization which captures how people
	combine these functions in a selected collection of spreadsheets, quantifying
	the broad patterns of function nesting while also being specific enough to find
	anomalous formulae and link them back to the specific cell in spreadsheets
	where they originate. \par

@@

	\section{Related Work} \label{related-work} Visualizations excel for their
	ability to communicate information quickly and
	efficiently~\cite{baeza1999modern}; the visualization of spreadsheets,
	therefore, has offered much room for research. Several previous tools work
	within a spreadsheet, seeking to clarify the connections between the visible
	values in cells, the formulae that produce them, and the dependencies between
	cells that inform them. Igarashi and colleague's fluid
	visualizations~\cite{igarashi1998fluid}, for example, do this by imposing
	lines, color, and animation over the spreadsheet to visually explain these
	tangled connections. Likewise, Clermont explored ways of grouping cells by
	color and border that were related through similar functions, neighbors, or
	references~\cite{clermont2003scalable}, and Hipfl extended this work by
	incorporating the layouts and labels of spreadsheets~\cite{hipfl2008using}.
	Other tools focus on creating visualizations external to the sheets. Hermans
	and colleagues address a spreadsheet programmer's information needs by making
	dataflow diagrams from individual sheets through the tools
	GyroSAT~\cite{hermans2011supporting} and Breviz~\cite{hermans2011breviz}.
	\toolname is among the latter class, creating standalone visualizations about
	spreadsheets instead of drawing connections inside the spreadsheet itself. In
	contrast to Hermans' work, our work focuses on representing functions and how
	they comprise complex formula, rather than exploring the data structures of any
	particular spreadsheet.  \par

	Many other tools inspect the structure of formulae but often from the
	perspective of finding bad design.  Following Fowler's initial description of
	code smells in object-oriented design~\cite{fowler2009refactoring}, many
	researchers have applied similar observations to spreadsheets and have
	cataloged classes of problematic
	formulae~\cite{hermans2012detecting}~\cite{cunha2012towards}~\cite{asavametha2012detecting}, often accompanying them with detection tools~\cite{abreu2014smelling} and investigations into real datasets~\cite{jansen2015code}. For example, Hermans, leveraging her work with the aforementioned dataflow diagrams, has worked with assessing common inter-worksheet smells, such as feature envy and inappropriate intimacy, wherein references between worksheets (rather than cells) in a file suggest a problem~\cite{hermans2012detectinginter}. Other tools, like Badam's Excel refactoring~\cite{badame2012refactoring}, are concerned more with understanding the formula in service of changing them. However, these studies define standards of poor design and orient their searches around them. Though \toolname supports the search for bad design, as we will see, the tool itself takes a much more agnostic approach in presentation, offering examples without regard for design quality. \par

	Implicit in this study is an exploration of the Excel API and how people use
	it. Previous research, such as the first study on the Enron
	spreadsheets~\cite{hermans2015enron} or in its comparison with the EUSUS
	corpus~\cite{jansen2015enron}, quantify how ofter users employ certain
	functions but not how they combine them. Researchers have applied such
	questions to other contexts as well, such as Murphy and colleague's study into
	how Java developers use Eclipse and what commands they use most
	often~\cite{murphy2006java}. Others focus on the documentation of the API
	itself: Robillard and DeLine, for example conducted a series of surveys and
	interviews to diagnose common problems with learning APIs, and they found that
	code examples, one of the focuses of this tool, is one of the most important
	aids to have in learning a new system~\cite{robillard2011field}. Though this
	paper applies the tool to the Enron corpus specifically, the tool itself is not
	limited to this; it can be readily applied to any collection of Excel
	spreadsheets a user gives it.

@@

	\subsection{Design Goals} \label{goals} In visualizing the spreadsheet data, we
	follow a few core goals for what the tool should accomplish. Furthermore, with
	these goals, we also considered the likely challenges of achieving those goals
	and particular outcomes which we considered beyond the scope of the research.
	We present both sides of these goals below:

@@

		\item [!2] \textbf{Do not overwhelm with numbers.} For popular functions
		especially, we anticipated that we'd have to show a lot of related data at
		once. Though everything should be accessible in accordance with Goal 1, we
		should design in a way that reduces noise.

@@

	The visualization so far comprises two types of nodes: the circle
	\includegraphics{glossary-greenonly}, which represents a discrete function in
	the formula and is size according to its frequency relative to the number of
	times to root function appeared; and the numbered squares
	\includegraphics{glossary-blue}, which represent the positions of arguments
	within its parent function. From this, we can confirm that, of the times it was
	observed, IF can have at most three arguments passed into it, which corresponds
	with its specifications in the
	API\footnote{https://support.office.com/en-us/article/IF-function-69aed7c9-4e8a-4755-a9bc-aa8bbff73be2}. \par

	Knowing that it is the first argument which contains the conditionals, we click
	on the square labeled ``1" to explore. To save space, when more than 10 unique
	arguments were found in a position, the tool displays only the first
	ten, with an option to display the rest by clicking the arrow
	\includegraphics{glossary-arrow}.

@@

		\item  \vspace{.25cm} \includegraphics{glossary-solidline} - \textbf{Solid
			lines} represent a function's acceptance of its parameters. As such, they
		connect the function nodes to argument nodes.

@@

		\item  \vspace{.25cm} \includegraphics[scale=.75]{glossary-lightblue} -
		\textbf{Optional argument trees} are variations of trees that show only
		formulae with a specific number of arguments. For example, the function AND
		can take any number of arguments, up to 255. By default, the tree shows data
		from AND functions with any number of argument, but these trees, marked with a
		darker blue color, can show how AND was used when \textit{only} two arguments
		were passed into it, or \textit{only} four, and so on. See
		Figure~\ref{fig:optional} for how the data between tree types differs.
	\end{itemize}

@@

		\item \textit{Presentation}: The JSON files, meanwhile, feed into the
		presentation code, implemented in Javascript with much help from the
		visualization library D3\footnote{https://d3js.org/}. We chose D3 primarily
		for its accessibility: given the JSON, the visualization can be simply
		embedded into a webpage accessible through a browser. \end{itemize} The code and a
	demonstration of \toolname can be found at
	https://github.com/DeveloperLiberationFront/Excel-Function-Visualizer.

@@

	However, because \toolname makes no judgments beforehand about which designs to
	emphasize as good or bad, the tool can be used to find strange or suboptimal
	designs that exist outside these obvious categories. Consider the second
	argument of the ``='' operator, already shown expanded in
	Figure~\ref{fig:fullpic}: among the top five argument types for the position in
	an IF function were boolean values. This corresponds to designs in which
	conditions already in a boolean form were then checked against "=TRUE" or
	"=FALSE", such as in the formula \textit{IF(Formulas!C77=TRUE, "Yes")}, which
	appears in one of the sheets. Though this practice does not break anything in
	itself, it nevertheless represents redundant operations in formula design and
	indicates a possible misunderstanding of how the IF function works over the
	hundreds of examples of this boolean comparison that our tool can return.

@@

	\begin{wrapfigure}{O}{.25\textwidth}
		\centering \includegraphics[width=.20\textwidth]{SUM}  \caption{An abundance of parameters in SUM.
			Encountering this, P4 remarked, "Woah."} \label{fig:sum} \end{wrapfigure}

	Additionally, because the files in the Enron set had name with the author as a
	prefix, P2 commented that they could use the tool in a company setting to
	follow up with repeat offenders. For instance, they first encountered a
	particular author's name when investigating an AVERAGE function with 18
	distinct arguments for which our participant could not discern any logical
	order, even after opening the source spreadsheet; the same name popped up again
	as P2 looked at a RATE function relying on several hard-coded values. P2 said
	that, if they were a manager or project leader, this recurrence was a sign to
	talk to the author personally and figure out why these problems were there, as
	to prevent more in the future. They also added that these problems would grow
	worse if this particular author left the company and no one remained who
	understood these formulae, a concern that aligns with Hermans' transfer
	scenarios \cite{hermans2011supporting}.

@@

	\subsection{Informative Qualification} As we found in the case study, the tool
	excels at making formula examples available throughout a tree. 	
	However, the benefits described in the previous sections imply a hidden
	complication: if the tool educates through example but does not explicitly
	distinguish between good design and bad design, how can we be sure that people
	will not inadvertently learn bad design from it? In the user study, we
	discussed its use in education and smell detection by bringing up the issues
	surrounding the LOOKUP functions; this discussion, though, presupposes a
	distinguishing eye that knew which examples practiced good design and which
	exhibited the common errors. A beginner might not know the difference.
	Robillard and Deline, in their study, remarked that good API examples should
	demonstrate best practices, but, right now, we have no way to assure this with
	\toolname. As with the problems in achieving unbiased exploration, correcting
	for this danger may entail a reevaluation of design goals.

@@

	\section{Conclusion} This paper presents a tool to support the exploration of
	function combinations throughout a given spreadsheet dataset. By taking an
	agnostic approach to design quality, it attempts to convey the dataset's
	spreadsheet practices as is: the common along with the anomalous, and the
	well-designed along with the odorous. As seen in the user and case studies,
	this allows it to take on several potential roles at once, such as an
	environment for search out bad code smells as well as a repository for examples
	for learning new techniques. \par

>>>>

	This paper contributes such a tool, \toolnameend, that attempts to fulfill this
	need for better exploration by building on the aforementioned spreadsheet
	corpora. Working with spreadsheets in Excel, \toolname focuses primarily on the
	functions, like SUM and IF and hundreds more, which are the building blocks of
	formulae that determine values across a spreadsheet. The tool
	creates an interactive, exploratory visualization which captures how people
	combine these functions in a selected collection of spreadsheets, quantifying
	the broad patterns of function nesting while also being specific enough to find
	anomalous formulae and link them back to the specific cell in spreadsheets
	where they originate. \par

@@

	\section{Related Work} \label{related-work} Visualizations excel for their
	ability to communicate information quickly and
	efficiently~\cite{baeza1999modern}; the visualization of spreadsheets,
	therefore, has offered much room for research. Several previous tools work
	within a spreadsheet, seeking to clarify the connections between the visible
	values in cells, the formulae that produce them, and the dependencies between
	cells that inform them. Igarashi and colleague's fluid
	visualizations~\cite{igarashi1998fluid}, for example, do this by imposing
	lines, color, and animation over the spreadsheet to visually explain these
	tangled connections. Likewise, Clermont explored ways of grouping cells by
	color and border that were related through similar functions, neighbors, or
	references~\cite{clermont2003scalable}, and Hipfl extended this work by
	incorporating the layouts and labels of spreadsheets~\cite{hipfl2008using}.
	Other tools focus on creating visualizations external to the sheets. Hermans
	and colleagues address a spreadsheet programmer's information needs by making
	dataflow diagrams from individual sheets through the tools
	GyroSAT~\cite{hermans2011supporting} and Breviz~\cite{hermans2011breviz}.
	\toolname is among the latter class, creating standalone visualizations about
	spreadsheets instead of drawing connections inside the spreadsheet itself. In
	contrast to the other visualization work, our work focuses on representing functions and how
	they combine into complex formula, rather than exploring the data structures of any
	particular spreadsheet.  \par

	Many other tools inspect the structure of formulae but often from the
	perspective of finding bad design.  Following Fowler's initial description of
	code smells in object-oriented design~\cite{fowler2009refactoring}, many
	researchers have applied similar observations to spreadsheets and have
	cataloged classes of problematic
	formulae~\cite{hermans2012detecting}~\cite{cunha2012towards}~\cite{asavametha2012detecting}, often accompanying them with detection tools~\cite{abreu2014smelling} and investigations into real datasets~\cite{jansen2015code}. For example, Hermans, leveraging her work with the aforementioned dataflow diagrams, has worked with assessing common inter-worksheet smells, such as feature envy and inappropriate intimacy, wherein references between worksheets (rather than cells) in a file suggest a problem~\cite{hermans2012detectinginter}. Other tools, like Badame's Excel refactoring~\cite{badame2012refactoring}, are concerned more with understanding the formula in service of changing them. However, these studies define standards of poor design and orient their searches around them. Though \toolname supports the search for bad design, as we will see, the tool itself takes a much more agnostic approach in presentation, offering examples without regard for design quality. \par

	Implicit in this study is an exploration of the Excel API and how people use
	it. Previous research, such as the first study on the Enron
	spreadsheets~\cite{hermans2015enron} or in its comparison with the EUSUS
	corpus~\cite{jansen2015enron}, quantifies how ofter users employ certain
	functions but not how they combine them. Researchers have applied such
	questions to other contexts as well, such as Murphy and colleague's study into
	how Java developers use Eclipse and what commands they use most
	often~\cite{murphy2006java}. Others focus on the documentation of the API
	itself: Robillard and DeLine, for example conducted a series of surveys and
	interviews to diagnose common problems with learning APIs, and they found that
	code examples, one of the focuses of this tool, is one of the most important
	aids to have in learning a new system~\cite{robillard2011field}. Though this
	paper applies the tool to the Enron corpus specifically, the tool itself is not
	limited to this; it can be readily applied to any collection of Excel
	spreadsheets a user gives it.

@@

	\subsection{Design Goals} \label{goals} In visualizing the spreadsheet data, we
	set out a few core goals for what the tool should accomplish. Furthermore, with
	these goals, we also considered the likely challenges of achieving those goals
	and particular outcomes which we considered beyond the scope of the research.
	We present both sides of these goals below:

@@

		\item [!2] \textbf{Do not overwhelm with numbers.} For popular functions
		especially, we anticipate that we'll have to show a lot of related data at
		once. Though everything should be accessible in accordance with Goal 1, we
		should design in a way that reduces noise.

@@

	The visualization so far comprises two types of nodes: the circle
	\includegraphics{glossary-greenonly}, which represents a discrete function in
	the formula and is size according to its frequency relative to the number of
	times to root function appeared; and the numbered squares
	\includegraphics{glossary-blue}, which represent the positions of arguments
	within its parent function. From this, we can confirm that, of the times it was
	observed, IF can have at most three arguments passed into it, which corresponds
	with its specifications in the
	API. \par

	Knowing that it is the first argument which contains the conditionals, we click
	on the square labeled ``1" to explore. To save space, when the tool finds more than 10 unique
	arguments in a position, it displays only the first
	ten, with an option to display the rest by clicking the arrow
	\includegraphics{glossary-arrow}.

@@

		\item  \vspace{.25cm} \includegraphics{glossary-solidline} - \textbf{Solid
			lines} represent a function's list of parameters. As such, they
		connect the function nodes to argument nodes.

@@

		\item  \vspace{.25cm} \includegraphics[scale=.75]{glossary-lightblue} -
		\textbf{Optional argument trees} are variations of trees that show only
		formulae with a specific number of arguments. For example, the function AND
		can take any number of arguments, up to 255. By default, the tree shows data
		from AND functions with any number of arguments, but these trees, which have argument nodes of a lighter blue color, can show how AND was used when \textit{only} two arguments
		were passed into it, or \textit{only} four, and so on. See
		Figure~\ref{fig:optional} for how the data between tree types differs.
	\end{itemize}

@@

		\item \textit{Presentation}: The JSON files, meanwhile, feed into the
		presentation code, implemented in Javascript with much help from the
		visualization library D3\footnote{https://d3js.org/}. We chose D3 primarily
		for its accessibility: given the JSON, the visualization can be simply
		embedded into a webpage accessible through a browser. \end{itemize} A
	demonstration of \toolname can be found at
	https://github.com/DeveloperLiberationFront/Excel-Function-Visualizer.

@@

	However, because \toolname makes no judgments beforehand about which designs to
	emphasize as good or bad, the tool can be used to find strange or suboptimal
	designs that exist outside these readily visible categories. Consider the second
	argument of the ``='' operator, already shown expanded in
	Figure~\ref{fig:fullpic}: among the top five argument types for the position in
	an IF function were boolean values. This corresponds to designs in which
	conditions already in a boolean form were then checked against "=TRUE" or
	"=FALSE", such as in the formula \textit{IF(Formulas!C77=TRUE, "Yes")}, which
	appears in one of the sheets. Though this practice does not break anything in
	itself, it nevertheless represents redundant operations in formula design and
	indicates a possible misunderstanding of how the IF function works over the
	hundreds of examples of this boolean comparison that our tool can return.

@@

	\begin{wrapfigure}{L}{.25\textwidth}
		\centering \includegraphics[width=.20\textwidth]{SUM}  \caption{An abundance of parameters in SUM, which was a commonly noted smell in the user study} \label{fig:sum} \end{wrapfigure}

	Additionally, because the files in the Enron set had name with the author as a
	prefix, P2 commented that they could use the tool in a company setting to
	follow up with repeat offenders. For instance, they first encountered a
	particular author's name when investigating an AVERAGE function with 18
	distinct arguments for which our participant could not discern any logical
	order even after opening the source spreadsheet; the same name popped up again
	as P2 looked at a RATE function relying on several hard-coded values. P2 said
	that, if they were a manager or project leader, this recurrence was a sign to
	talk to the author personally and figure out why these problems were there, as
	to prevent more in the future. They also added that these problems would grow
	worse if this particular author left the company and no one remained who
	understood these formulae, a concern that aligns with Hermans' transfer
	scenarios \cite{hermans2011supporting}.

@@

	\subsection{Informative Qualification} As we found in the case study, the tool
	excels at making formula examples available throughout a tree. 	
	However, the benefits described in the previous sections imply a hidden
	complication: if the tool educates through example but does not explicitly
	distinguish between good design and bad design, how can we be sure that people
	will not inadvertently learn bad design from it? In the user study, we
	discussed its use in education and smell detection by bringing up the issues
	surrounding the LOOKUP functions; this discussion, though, presupposes a
	distinguishing eye that knew which examples practiced good design and which
	exhibited the common errors. A beginner might not know the difference.
	Robillard and Deline, in their study, remarked that good API examples should
	demonstrate best practices, but, right now, we have no way to assure this with
	\toolnameend. As with the problems in achieving unbiased exploration, correcting
	for this danger may entail a reevaluation of design goals.

@@

	\section{Conclusion} This paper presents a tool to support the exploration of
	function combinations throughout a given spreadsheet dataset. By taking an
	agnostic approach to design quality, it attempts to convey the dataset's
	spreadsheet practices as is: the common along with the anomalous, and the
	well-designed along with the odorous. As seen in the user and case studies,
	this allows it to take on several potential roles at once, such as an
	environment for investigating code smells as well as a repository for examples
	for learning new techniques. \par
